Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-27 paper.tex
Edits on the Starvation/Deadlocking Section

\usepackage{xcolor}

\pagecolor[rgb]{0.5,0.5,0.5}
\color[rgb]{1,1,1}

@@

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{Forward-Progress Guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward progress guarantee}; without this guarantee, it risks indefinite starvation.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward progress guarantees (FPG). Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL). Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a \emph{forward progress guarantee} of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{Stream Scan}.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Why does \emph{Chained Scan} Rely on Forward-Progress Guarantees?}
A \emph{forward-progress guarantee} (FPG) ensures that all occupant subgroups (of possibly different workgroups) eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically defined and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their findings demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior assumptions~\cite{sorensen2018}.

This has direct implications for \emph{Chained Scan}, which was \emph{previously} classified as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. Under HSA, this structure guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then any dependent subgroup may spin indefinitely, waiting on an update that never arrives.

@@

\begin{table}
  \small
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Algorithm         & LOBE & OBE & HSA & Unfair \\
    \midrule
    CS (Workgroup ID) & Yes  & No  & Yes & No     \\
    CS (Atomic ID)    & Yes  & Yes & No  & No     \\
    Ours              & Yes  & Yes & Yes & Yes    \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of Chained Scan (CS) Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
\end{table}

\thomas{Is the one-to-one workgroup to multiprocessor rebuttal still neccessary here?}

\subsection{Why is reliance on Forward Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

Another potential bottleneck is the speed of atomic operations. On some older GPUs—such as pre-Kepler NVIDIA architectures—atomics do not operate within the cache, resulting in significantly higher latency. Slower atomics lead to delayed tile updates, which in turn increase lookback lengths. While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.

>>>>

@@

\begin{abstract}
  We present \emph{Decoupled Fallback}, a method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding deadlock. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled-Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WGPU and realizes the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled-Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\section{Introduction}
Scan is one of the most ubiquitous operations in GPGPU computing, with applications in sorting~\cite{adinets2022onesweepfastersignificantdigit}, sparse matrix operations~\cite{BellGarland2009}, and more. The current state-of-the-art prefix scan algorithm, Merrill and Garland's \emph{Decoupled Lookback}, achieves near-ideal performance on NVIDIA hardware via CUDA by leveraging the \emph{Chained Scan} architecture. This approach hybridizes serial and parallel strategies to read and write each element exactly once, achieving the theoretical minimum global memory movement of $O(2n)$. \emph{Chained Scans} that fully saturate device memory bandwidth reach the \emph{Speed of Light} (SOL) performance limit---the theoretical maximum throughput constrained by the physical limits of the device. However, \emph{Decoupled Lookback} relies on architectural and language-specific guarantees provided by the NVIDIA ecosystem, most notably a device scheduler property known as a \emph{forward-progress guarantee}; without this guarantee, it risks indefinite starvation.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPGPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers—most notably Apple's M series and ARM—lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

This work is guided by two main goals. (1) \textbf{Portability:} We aim to develop a scan implementation that retains the benefits of the \emph{Chained Scan} architecture while being suitable for a diverse range of hardware vendors and architectures, including those without forward-progress guarantees (FPG) or support for 64-bit atomics. \textbf{Performance:} We aim for near speed-of-light execution to the greatest extent possible allowed by the underlying hardware and programming model.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving speed-of-light performance without requiring forward-progress guarantees (FPG). Our implementation is designed for compatibility with shading languages, making it suitable for modern graphics APIs and implemented in the WebGPU Shading Language (WGSL). Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

In any \emph{Chained Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a forward-progress guarantee of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

Although \emph{Decoupled Lookback} increases global memory access during the propagation phase to $O(o(n/t))$, the serial ordering of work tile processing confines spine memory accesses to a roughly $o$-width sliding window, making them highly likely to be cache-resident. As a result, the additional global memory traffic along the spine introduces negligible overhead. With total global memory movement of $O(2n+o(n/t))= O(2n)$, \emph{Decoupled Lookback} achieves full utilization of device memory bandwidth and delivers speed-of-light performance without the work tile size tuning required by \emph{StreamScan}.

@@

Given these constraints, oversubscription and \emph{Chained Scan} emerge as the more portable and preferred architecture. Furthermore, its single-kernel design and serial ordering make it suitable for \emph{in-situ} operations, benefiting applications such as data deduplication, sparse data compaction, error correction, and interval merging. However, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that ARM and Apple’s GPUs do not. Without forward-progress guarantees, \emph{Chained Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

\subsection{Progress Guarantees and Their Impact on \emph{Chained Scan}}
A \emph{forward-progress guarantee} ensures that all occupant subgroups---of possibly different workgroups---eventually make progress toward termination, preventing indefinite starvation. Prior to Sorensen et al.~\cite{sorensen2018,sorensen2021}, developers largely relied on untested, folklore-like assumptions about GPU progress guarantees. Instead, Sorensen et al.\ systematically proposed and tested three progress models:\footnote{Note, a ``workgroup'' in Sorensen et al.'s definition is not necessarily in-flight and occupant on a multiprocessor.}
\begin{itemize}
  \item \textbf{Occupancy-Bound Execution (OBE)}: Workgroups that have made at least one execution step.
  \item \textbf{Heterogeneous System Architecture~\cite{} (HSA)}: The workgroup with the lowest ID that has not yet terminated.
  \item \textbf{Linear OBE (LOBE)}: Workgroups that have made at least one step, and any workgroup with a lower ID than a workgroup that has made at least one step.
\end{itemize}
Their findings in in \cite{sorensen2021} demonstrated that Apple and ARM GPUs do not support LOBE, contradicting their prior hypothesis in \cite{sorensen2018}.

Sorensen et al. classify \emph{Chained Scan} as a \emph{one-way producer-consumer} (PC) HSA algorithm: a workgroup only consumes from lower-ID counterparts, ensuring that execution proceeds in a strictly ordered fashion. HSA guarantees termination, as the lowest ID workgroup is always given forward progress. If subgroups from the lowest ID workgroup are not guaranteed execution, then dependent workgroups may spin indefinitely, waiting on an update that never arrives.

@@

Thus, by dynamically acquiring IDs through atomics, \emph{Chained Scan} shifts its progress model requirement from HSA to OBE. While LOBE encompasses both HSA and OBE, its stricter ordering guarantees are unnecessary for correctness. Ultimately, regardless of a device’s specific progress model, any architecture lacking forward-progress guarantees will fail to execute \emph{Chained Scan}. In short:

\begin{table}[h!]
  \small
  \centering
  \begin{tabular}{lcccc}
    \toprule
    Algorithm         & LOBE & OBE & HSA & None \\
    \midrule
    \emph{Chained Scan} with Workgroup ID & Yes  & No  & Yes & No     \\
    \emph{Chained Scan} with Atomic ID    & Yes  & Yes & No  & No     \\
    Ours                                  & Yes  & Yes & Yes & Yes    \\
    \bottomrule
  \end{tabular}
  \caption{Comparison of \emph{Chained Scan} Variants and Our Implementation Across Progress Models.\label{tab:progress_models}}
\end{table}

\subsection{Why is reliance on Forward-Progress Guarantees a Portability Problem?}
The absence of FPG in key hardware segments poses a critical challenge for developers. Our testing shows that, at best, running \emph{Chained Scan} without FPG results in severe performance degradation, and at worst, risks \emph{Timeout Device Recovery} and program crashes.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{graphics/Figure_1.pdf}
  \caption{Execution times of \emph{Decoupled Lookback (PLACEHOLDER DATA)}}
\end{figure}
Exacerbating this issue, no major graphics API provides a mechanism to query FPG support. This lack of transparency forces developers to design algorithms without knowing whether critical guarantees are available on the target hardware. In shader environments, where cross-platform portability is essential, this uncertainty compels a fallback to older, less efficient strategies like \emph{Reduce-then-Scan}, sacrificing performance for broader compatibility.

@@

Another potential bottleneck is the speed of atomic operations. On some older GPUs---such as pre-Fermi NVIDIA architectures---atomics do not operate within the cache, resulting in significantly higher latency. Slower atomics lead to delayed tile updates, which in turn increase lookback lengths. While our simulated deadlocks provide an artificial means of stress-testing execution, real deadlocks manifest as an effective \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
